---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# nlfh

`nlfh` fits linear and nonlinear Bayesian Fay-Herriot models for small area
estimation with area-level direct estimates and known sampling variances.

The main entry point is `fit_fh()`. Use `method = "linear"` for the standard
linear Fay-Herriot model, `method = "rnn"` for the random-weight neural network
extension described by Parker (2024), and `method = "bart"` for the BART-FH
model described by Parker and Eideh (2026).

## Installation

You can install the development version of nlfh from GitHub with:

``` r
# install.packages("pak")
pak::pak("paparker/nlfh")
```

## Example

The included `acs_dat` data set contains 2021 direct estimates of median income
(`MedInc`) for the state of Missouri, its standard error (`MedIncSE`), and area-level covariates. Pass the
sampling variance separately as `MedIncSE^2`.

```{r example}
library(nlfh)

data(acs_dat)
acs_small <- as.data.frame(acs_dat[1:500, ])
example_control <- list(n_iter = 500, burn_in = 250, progress = FALSE)

fit_linear <- fit_fh(
  MedInc ~ SNAPRate + PovRate + White + Black + Hispanic + Asian,
  sampling_variance = MedIncSE^2,
  data = acs_small,
  method = "linear",
  control = example_control
)

summary(fit_linear)
```

The formula interface specifies the available predictors. For nonlinear methods,
it does not impose an additive linear mean structure; the model estimates an
unknown function of the model matrix.
By default, the RNN model centers and scales non-intercept covariates before
fitting; use `control = list(scale = FALSE)` to disable this covariate
preprocessing. The RNN response and sampling variances are standardized
internally and returned on the original response scale.

```{r nonlinear-example}
fit_rnn <- fit_fh(
  MedInc ~ .,
  sampling_variance = MedIncSE^2,
  data = acs_small,
  method = "rnn",
  control = example_control
)

fit_bart <- fit_fh(
  MedInc ~ SNAPRate + PovRate + White,
  sampling_variance = MedIncSE^2,
  data = acs_small,
  method = "bart",
  control = example_control
)

c(linear = fit_linear$dic, rnn = fit_rnn$dic, bart = fit_bart$dic)
fit_bart$variable_importance
```

For BART fits, the first model-matrix column is treated as a
baseline/intercept column and is excluded from `variable_importance`. With the
formula interface this is usually the default `(Intercept)` column; with the
matrix interface, put the baseline or intercept column first.

You can also use the matrix interface directly.

```{r matrix-example}
X <- model.matrix(
  MedInc ~ SNAPRate + PovRate + White + Black + Hispanic + Asian,
  data = acs_small
)

fit_matrix <- fit_fh(
  y = acs_small$MedInc,
  X = X,
  sampling_variance = acs_small$MedIncSE^2,
  method = "linear",
  control = example_control
)

dim(fit_matrix$predictions)
```

## References

Parker, P. A. (2024). Nonlinear Fay-Herriot Models for Small Area Estimation
Using Random Weight Neural Networks. *Journal of Official Statistics*, 40(2),
317-332. doi:10.1177/0282423X241244671

Parker, P. A. and Eideh, A. (2026). BART-FH: Flexible Nonlinear Modeling for
Small Area Estimation. *Journal of Survey Statistics and Methodology*, 00,
1-18. doi:10.1093/jssam/smaf050
