---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# nlfh

`nlfh` fits linear and nonlinear Bayesian Fay-Herriot models for small area
estimation with area-level direct estimates and known sampling variances.

The main entry point is `fit_fh()`. Use `method = "linear"` for the standard
linear Fay-Herriot model, `method = "rnn"` for the random-weight neural network
extension, and `method = "bart"` for the BART-FH model.

## Installation

You can install the development version of nlfh from GitHub with:

``` r
# install.packages("pak")
pak::pak("paparker/nlfh")
```

## Example

The included `acs_dat` data set contains a direct estimate of median income
(`MedInc`), its standard error (`MedIncSE`), and area-level covariates. Pass the
sampling variance separately as `MedIncSE^2`.

```{r example}
library(nlfh)

data(acs_dat)
acs_small <- as.data.frame(acs_dat[1:500, ])
example_control <- list(n_iter = 500, burn_in = 250, progress = FALSE)

fit_linear <- fit_fh(
  MedInc ~ SNAPRate + PovRate + White + Black + Hispanic + Asian,
  sampling_variance = MedIncSE^2,
  data = acs_small,
  method = "linear",
  control = example_control
)

summary(fit_linear)
```

The formula interface specifies the available predictors. For nonlinear methods,
it does not impose an additive linear mean structure; the model estimates an
unknown function of the model matrix.

```{r nonlinear-example}
fit_rnn <- fit_fh(
  MedInc ~ .,
  sampling_variance = MedIncSE^2,
  data = acs_small,
  method = "rnn",
  control = example_control
)

fit_bart <- fit_fh(
  MedInc ~ SNAPRate + PovRate + White,
  sampling_variance = MedIncSE^2,
  data = acs_small,
  method = "bart",
  control = example_control
)

c(linear = fit_linear$dic, rnn = fit_rnn$dic, bart = fit_bart$dic)
```

You can also use the matrix interface directly.

```{r matrix-example}
X <- model.matrix(
  MedInc ~ SNAPRate + PovRate + White + Black + Hispanic + Asian,
  data = acs_small
)

fit_matrix <- fit_fh(
  y = acs_small$MedInc,
  X = X,
  sampling_variance = acs_small$MedIncSE^2,
  method = "linear",
  control = example_control
)

dim(fit_matrix$predictions)
```
